Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
Trimming

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of
GD-1, for the first time we find evidence of stream members off of the main
track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Stellar streams are formed during the tidal disruption of stellar systems by the
gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Dynamically-cold stellar streams, formed from disrupted globular clusters, can
also retain imprints on their phase-space density from encounters with a massive
perturber for close to their total survival time \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

@@

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
dark matter-only simulations \citep[scaled to the Milky Way halo mass and
reduced in number by a factor of 3 to account for disruption by the Galactic
disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
have up to one significant, wide ($\approx 5$--$7^\circ$) gap caused by an
interaction with a $10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.

Photometric studies of GD-1 revealed the presence of density variations and gaps
along the stream, as well as wiggles in the main stream track
\citep[][]{Carlberg:2013, DeBoer:2018}.
However, purely photometric studies of streams suffer from contamination of the
Milky Way foreground, whose small-scale spatial variations could partially
account for density inhomogeneities reported along cold streams
\citep[e.g.,][]{Ibata:2016}.
In this \article, we improve upon the selection of likely GD-1 members by using
astrometric data from the \gaia\ mission, in addition to precise photometry from
the \pans\ survey (\S\,\ref{sec:data}).
These data enable us to measure density variations along GD-1 to an
unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures
of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
Implications of this revolutionary remapping of GD-1 are discussed in
\S\,\ref{sec:discussion}.
% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
% (\citealt{Carlberg:2013}).
% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
% [summarize their conclusions].
% \todo{these analyses relied on binned data, still contaminated by MW stars}
%
% In this \article, we present clearest view of the GD-1 stream to date using
% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
% precise photometry from the \pans\ survey.
% We find XX under-densities gaps, some consistent with ...
% We revisit...do not try to model or fit for Galactic potential.
% Focus on stream properties, data.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the
overdensity of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
\phi_1 \lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to clean the
sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
Several of the under-densities and gaps hinted at from photometric selection
(\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in this
significantly cleaner map of the GD-1 stream.
Finally, at least two new features are visible near the stream: (1) ``the
spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
\lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion components
extracted in this way (dark lines), with the associated empirical scatters
(shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.
The median proper motion profile along the stream (panels (d)--(e)) are
consistent with previously measured mean proper motions from SDSS-USNOB
(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
individual stars (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
The median $V$-band extinction over the stream track is
$\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
$\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), promising a robust
estimate of the proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The observed density variations are therefore likely real morphological changes
along the GD-1 stream, as have been suggested before photometric selection alone
(\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by simulating the orbital evolution of cluster
stars once they are tidally stripped from the progenitor.
We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
and assume a uniform mass-loss history to generate the model stream, which
matches results from more detailed ($N$-body) numerical experiments of globular
cluster disruption \citep[e.g.,][]{Kupper:2012}.
% During simulation, star particles are released from the Lagrange points of the
% cluster with a spread in position and velocity set by the mass of the
% satellite and the location within the host galaxy, with some tunable scale
% parameters (using the parametrization and adopted scale parameters of
% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
% the resulting stream, but the density of stars along the stream depends in
% detail on the mass-loss history and internal kinematics of the progenitor
% system.  Here we assume a constant mass-loss rate, and release stream
% particles uniformly in time from each Lagrange point; these choices match
% results from more detailed numerical experiments of globular cluster
% disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed, three-component potential model
to represent the Milky Way, consisting of a disk (\citealt{Miyamoto:1975}), a
bulge (\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve between with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.

@@

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
then forward-generate the model stream using the method discussed above, and
release stream particles at each timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

\subsection{Off-track features}
\label{sec:res_gap}
In this section we test whether the overdensities apparent beyond the main stream track (Figure~\ref{fig:selection}) are truly associated with the GD-1 stream.
% this would be a novel insight into disruption from the \gaia\ mission
Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
To assess the origin of these features, we compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.

@@

For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
In both rows, arranged from left to right are the stream, the feature, and the control field.
Stars selected from the stream fields show a clear overdensity in the CMD selection box and they cluster in proper motions.
In detail, the two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.
On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$ in apparent length, reveals a
plausible location of the progenitor, and shows significant density variations
along the stream in unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of
GD-1, for the first time we find evidence of stream members off of the main
track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Stellar streams are formed during the tidal disruption of stellar systems by the
gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Dynamically-cold stellar streams formed from disrupted globular clusters can
also retain imprints on their phase-space density from encounters with a massive
perturber for long times \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

@@

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
% reduced in number by a factor of 3 to account for disruption by the Galactic
% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
dark matter-only simulations \citep{Springel:2008, Diemand:2008},
\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform the nature of dark matter \citep[e.g.][]{Bullock:2017}.

Photometric studies of GD-1 revealed the presence of density variations and gaps
along the stream, as well as wiggles in the main stream track
\citep[][]{Carlberg:2013, DeBoer:2018}.
However, purely photometric studies of streams suffer from contamination of the
Milky Way foreground, which, on small scales, could partially account for
density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
In this \article, we improve upon the selection of likely GD-1 members by using
astrometric data from the \gaia\ mission, combined with precise photometry from
the \pans\ survey (\S\,\ref{sec:data}).
These data enable us to measure density variations along GD-1 to an
unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures
of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
Implications of this revolutionary remapping of GD-1 are discussed in
\S\,\ref{sec:discussion}.
% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
% (\citealt{Carlberg:2013}).
% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
% [summarize their conclusions].
% \todo{these analyses relied on binned data, still contaminated by MW stars}
%
% In this \article, we present clearest view of the GD-1 stream to date using
% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
% precise photometry from the \pans\ survey.
% We find XX under-densities gaps, some consistent with ...
% We revisit...do not try to model or fit for Galactic potential.
% Focus on stream properties, data.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
batches of stars that have small latitude in the heliocentric GD-1 coordinate
system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars that pass the
selection, plotted in the GD-1 coordinate system.
The stream is already identifiable as the overdensity of stars around $\phi_2 =
0$ between longitudes $-60^\circ \lesssim \phi_1 \lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog.% and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
Several of the under-densities and gaps hinted at from photometric selection
(\citealt{Carlberg:2013, DeBoer:2018}) appear as striking features in this
significantly cleaner map of the GD-1 stream.
% Finally, at least two new features are visible near the stream: (1) ``the
% spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
% \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
% stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
% We discuss each of these in more detail in \sectionname
% s~\ref{sec:results}--\ref{sec:discussion} below.
Finally, at least two new features are visible near the stream: (1) ``the
spur,'' stars above the main stream track (in $\phi_2$), and (2) ``the blob,''
stars below the main stream track, both highlighted in
\figurename~\ref{fig:selection}.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion components
extracted in this way (dark lines), with the associated empirical scatters
(shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.
The median proper motion profiles along the stream (panels (d)--(e)) are
consistent with previous mean proper motion measurements
(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
individual stars (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), promising a robust
estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The observed density variations are therefore likely real morphological changes
along the GD-1 stream, as have been suggested before photometric selection alone
(\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material.
% that has not yet phase-mixed in multiple dimensions.

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by simulating the orbital evolution of globular
cluster stars once they are tidally stripped from the progenitor.
We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
and assume a uniform mass-loss history to generate the model stream \citep[e.g.,][]{Kupper:2012}.
%, which
% matches results from more detailed ($N$-body) numerical experiments of globular
% cluster disruption \citep[e.g.,][]{Kupper:2012}.
% During simulation, star particles are released from the Lagrange points of the
% cluster with a spread in position and velocity set by the mass of the
% satellite and the location within the host galaxy, with some tunable scale
% parameters (using the parametrization and adopted scale parameters of
% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
% the resulting stream, but the density of stars along the stream depends in
% detail on the mass-loss history and internal kinematics of the progenitor
% system.  Here we assume a constant mass-loss rate, and release stream
% particles uniformly in time from each Lagrange point; these choices match
% results from more detailed numerical experiments of globular cluster
% disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed Milky Way model consisting of a
disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
spherical dark matter halo (\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
%use L-BFGS-B optimization as implemented in the \package{scipy}
% \texttt{Python} package to
We maximize the likelihood of the orbit given sky track and proper motions from
this work, with mean distance and giant star radial velocities from
\cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison stream model. % of the stream properties.

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

\subsection{Off-track features}
\label{sec:res_gap}
In this section we test whether the overdensities apparent beyond the main stream track (Figure~\ref{fig:selection}) are truly associated with the GD-1 stream.
% this would be a novel insight into disruption from the \gaia\ mission
Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
We compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.

@@

For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
% In both rows, arranged from left to right are the stream, the feature, and the control field.
Stars selected from the stream fields show a clear overdensity in the CMD selection box and cluster in proper motions.
The two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.
On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.
